```{r}
#Assessment module 11

#Load packages
library(tidymodels)
library(ggplot2)
library(here)
library(glmnet)
library(ranger)

#Set seed
rngseed = 1234
set.seed(rngseed)
```


```{r}
#Load data
data <- readRDS(here("ml-models-exercise/", "data_clean.rds"))

#Race category adjustment
data <- data %>%
  mutate(RACE = case_when(
    RACE == 1 ~ 1,
    RACE == 2 ~ 2,
    RACE %in% c(7, 88) ~ 3,
  ))

#Race possible categories bases on manuscript
data <- data %>%
  mutate(RACE_cat = case_when(
    RACE == 1 ~ "caucasian",
    RACE == 2 ~ "non-caucasian",
    RACE == 3 ~ "undisclosed"
  ))

```

```{r}
#Correlation plots for continuous variables
data_continuous <- data.frame("Y" = data$Y, 
#"DOSE" = data$DOSE, dose variable removed
"WT" = data$WT, 
"HT" = data$HT
)

#Pairwise comparison
pairs(data_continuous)

#Correlation matrix
cor(data_continuous)

#New feature BMI
data$BMI <- (data$WT/(data$HT^2))
```

```{r}
#Creating 3 models

#Linear regression
model1 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ DOSE + AGE + SEX + WT + HT, data = data)

#LASSO
model2 <- linear_reg(penalty = 0.1) %>%
  set_engine("glmnet") %>% 
  fit(Y ~ DOSE + AGE + SEX + WT + HT, data = data)

#Random forest
model3 <- rand_forest(mode = "regression") %>% 
  set_engine("ranger", seed = rngseed) %>% 
  fit(Y ~ DOSE + AGE + SEX + WT + HT, data = data)

```

```{r}
#Predictions for each model
pred1 <- predict(model1, new_data = data) 
pred2 <- predict(model2, new_data = data)
pred3 <- predict(model3, new_data = data) 

#Adding predictions into data
data <- data %>%
  mutate(
    pred1 = predict(model1, new_data = data, type = "numeric")$.pred,
    pred2 = predict(model2, new_data = data, type = "numeric")$.pred,
    pred3 = predict(model3, new_data = data, type = "numeric")$.pred
  )

#RSME values for each model
rmse(data, truth = Y, estimate = pred1)
rmse(data, truth = Y, estimate = pred2)
rmse(data, truth = Y, estimate = pred3)

```

```{r}
#Plotting predicted vs observed

#Model1
ggplot(data, aes(y = Y, x = pred1)) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, color = "gray50", linewidth = 0.8, linetype = "dashed") +
    lims(x = c(0, 6000), y = c(0, 6000)) +
    theme_bw()

#Model2
ggplot(data, aes(y = Y, x = pred2)) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, color = "gray50", linewidth = 0.8, linetype = "dashed") +
    lims(x = c(0, 6000), y = c(0, 6000)) +
    theme_bw()

#Model3
ggplot(data, aes(y = Y, x = pred3)) + 
    geom_point() + 
    geom_abline(slope = 1, intercept = 0, color = "gray50", linewidth = 0.8, linetype = "dashed") +
    lims(x = c(0, 6000), y = c(0, 6000)) +
    theme_bw()
```


```{r}
#LASSO model tuning

#set specs
lasso_model <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

#set recipe
lasso_recipe <- recipe(Y ~ DOSE + AGE + SEX + WT + HT, data = data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors(), -all_outcomes())

#establish workflow using specs and recipe
lasso_workflow <- workflow() %>%
  add_model(lasso_model) %>%
  add_recipe(lasso_recipe)

#create penalty grid
lasso_grid <- tibble(
  penalty = 10^seq(-5, 2, length.out = 50)
)

#resampling with apparent()
lasso_app <- apparent(data)

#reset seed
set.seed(rngseed)

#Lasso grid tune object
lasso_tune <- tune_grid(
  lasso_workflow,
  resamples = lasso_app,
  grid = lasso_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

#print predictions
collect_predictions(lasso_tune)

#plot manually, autoplot() did not work
lasso_tune %>%
    collect_metrics(summarize = FALSE) %>%
    filter(.metric == "rmse") %>%
    ggplot(aes(x = log(penalty, base = 10), y = .estimate)) +
    geom_line() + 
    geom_point() +
    labs(
        x = "Penalty",
        y = "RMSE"
    )+
    theme_bw()
```

As the penalty increases, LASSO shrinks coefficients and reduces model flexibility, introducing bias and increasing RMSE; since linear regression is the unpenalized optimum, LASSO cannot achieve a lower RMSE than the linear model when evaluated on the training data.

```{r}
#tuning RF model

#model specs
rf_model <- rand_forest(
  mode = "regression",
  mtry = tune(),
  min_n = tune(),
  trees = 300
) %>%
  set_engine("ranger", seed = rngseed)

#model workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(lasso_recipe)  # reuse recipe

#model grid
rf_grid <- grid_regular(
  mtry(range = c(1, 7)),
  min_n(range = c(1, 21)),
  levels = 7
)

#RF tuning
rf_tune <- tune_grid(
  rf_workflow,
  resamples = apparent(data),
  grid = rf_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

#print predictions
collect_predictions(rf_tune)

#Collecting RMSE from rf tune
rf_results <- rf_tune %>%
  collect_metrics(summarize = FALSE) %>%
  filter(.metric == "rmse")

#plot manually, autoplot() did not work
ggplot(rf_results, aes(x = as.factor(mtry), y = as.factor(min_n), fill = .estimate)) +
  geom_tile() +
  labs(
    x = "mtry",
    y = "min_n",
    fill = "RMSE"
  ) +
  theme_bw()
```

### Repeating with CV

```{r}
# Set the seed 
set.seed(rngseed)

# Create the 5-fold CV, repeated 5 times
resamples_cv <- vfold_cv(data, v = 5, repeats = 5)
```

```{r}
#LASSO model tuning with CV

#set specs
lasso_model_cv <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

#set recipe
lasso_recipe_cv <- recipe(Y ~ DOSE + AGE + SEX + WT + HT, data = data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors(), -all_outcomes())

#establish workflow using specs and recipe
lasso_workflow_cv <- workflow() %>%
  add_model(lasso_model_cv) %>%
  add_recipe(lasso_recipe_cv)

#create penalty grid
lasso_grid_cv <- tibble(
  penalty = 10^seq(-5, 2, length.out = 50)
)

#reset seed
set.seed(rngseed)

#Lasso grid tune object
lasso_tune_cv <- tune_grid(
  lasso_workflow_cv,
  resamples = resamples_cv,
  grid = lasso_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

#print predictions
collect_predictions(lasso_tune_cv)

#plotting with autoplot
lasso_tune_cv %>% autoplot() + theme_bw()
```


```{r}
#tuning RF model with CV

#model specs
rf_model_cv <- rand_forest(
  mode = "regression",
  mtry = tune(),
  min_n = tune(),
  trees = 300
) %>%
  set_engine("ranger", seed = rngseed)

#model workflow
rf_workflow_cv <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(lasso_recipe)  # reuse recipe

#model grid
rf_grid_cv <- grid_regular(
  mtry(range = c(1, 7)),
  min_n(range = c(1, 21)),
  levels = 7
)

#RF tuning
rf_tune_cv <- tune_grid(
  rf_workflow,
  resamples = resamples_cv,
  grid = rf_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

#print predictions
collect_predictions(rf_tune_cv)

#Collecting RMSE from rf tune
rf_results_cv <- rf_tune_cv %>%
  collect_metrics(summarize = FALSE) %>%
  filter(.metric == "rmse")

#plotting with autoplot
rf_tune_cv %>% autoplot() + theme_bw()

```

When we switch to cross-validation, the RMSE for both models goes up. Without CV. Cross-validation gives us a much more realistic picture by repeatedly training the model on different slices of the data and testing it on the parts it hasn’t seen. This process  penalizes overfitting and shows us how well each model will actually perform on new, unseen data.