#Fitting exercise

```{r}
# Loading data
library(here)
data <- read.csv(here("fitting-exercise/", "Mavoglurant_A2121_nmpk.csv"))

str(data)

# Plotting DV ~ TIME
library(ggplot2)
ggplot(data, aes(y = DV, x = TIME)) + geom_point() + theme_bw()

ggplot(data, aes(y = DV, x = TIME, color = as.factor(DOSE))) + geom_point() + theme_bw()

ggplot(data, aes(y = DV, x = TIME, color = as.factor(DOSE))) + geom_point() + facet_wrap(~ DOSE) + theme_bw() + theme(strip.text = element_blank())
```

```{r}
# Removing OCC = 2
data_occ1 <- subset(data, OCC == 1)

# Removing time = 0
data_timefilter <-  subset(data_occ1, !(TIME == 0))

# Data only with time zero
data_timezero <-  subset(data_occ1, (TIME == 0))

library(dplyr)
# Summarizing - ChatGPT used for this code
sum_dv <- data_occ1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE))

# Merging datasets
merged_data <- left_join(data_timezero, sum_dv, by = "ID")
```

```{r}
# "Last" data cleaning
data_clean <- data.frame(Y  = merged_data$Y,
                        DOSE = merged_data$DOSE,
                        AGE = merged_data$AGE,
                        SEX = as.factor(merged_data$SEX),
                        RACE = as.factor(merged_data$RACE),
                        WT = merged_data$WT,
                        HT = merged_data$HT)

library(readr)
write_csv(data_clean, here("fitting-exercise/", "data_clean.csv"))

saveRDS(data_clean, here("ml-models-exercise/", "data_clean.rds"))
```

```{r}
# Plotting distributions for each one of the variables
hist(data_clean$Y) # Normally distirbuted, slight right skew
hist(data_clean$DOSE) # 3 doses used, will move to as.factor for certain plots
hist(data_clean$AGE) # few observation in 30-35 bin 
plot(data_clean$SEX) # sex 1 overrepresented
plot(data_clean$RACE) # Race 7 underrepresented
hist(data_clean$WT) # Normally distributed
hist(data_clean$HT) # Normally distributed
```

```{r}
# First round of plots
ggplot(data_clean, aes(y = Y, x = as.factor(DOSE))) + geom_boxplot() + theme_bw() # Hihgher dose, higher Y

ggplot(data_clean, aes(y = Y, x = AGE)) + geom_point() + theme_bw() # No clear pattern observed

ggplot(data_clean, aes(y = Y, x = SEX)) + geom_boxplot() + theme_bw() # Similar distributuion of SEX 1 and 2

ggplot(data_clean, aes(y = Y, x = RACE)) + geom_boxplot() + theme_bw() #Similar distribution for RACE 1, 2, 7, and 88

ggplot(data_clean, aes(y = Y, x = WT)) + geom_point() + theme_bw() # No clear pattern observed

ggplot(data_clean, aes(y = Y, x = HT)) + geom_point() + theme_bw() # No clear pattern observed

```

```{r}
# Second round of plots
ggplot(data_clean, aes(y = Y, x = as.factor(DOSE), levels =  SEX, fill = SEX)) + geom_boxplot() + theme_bw() # DOSE = 37.5 has wider distribution for SEX = 1

ggplot(data_clean, aes(y = Y, x = as.factor(DOSE), levels =  RACE, fill = RACE)) + geom_boxplot() + theme_bw() # DOSE = 37.5 has no representation of RACE = 7 and 88

ggplot(data_clean, aes(y = Y, x = WT, color = SEX)) + geom_point(size = 3) + theme_bw() + facet_wrap(~ SEX) + theme(strip.text = element_blank()) # SEX = 2 underrepresented as mentioned before, weight is less than SEX = 1

ggplot(data_clean, aes(y = Y, x = HT, color = RACE)) + geom_point(size = 3) + theme_bw() + facet_wrap(~ RACE) + theme(strip.text = element_blank()) # Only two observations of RACE = 7, very few observations of RACE = 88

ggplot(data_clean, aes(y = Y, x = WT, color = AGE, size = AGE)) + geom_point(alpha = 0.75) + theme_bw() # No clear pattern observed

ggplot(data_clean, aes(y = Y, x = HT, color = AGE, size = AGE)) + geom_point(alpha = 0.75) + theme_bw() # No clear pattern observed

```

```{r}
library(tidymodels)
library(dotwhisker)

#Loading engine for lm
linear_reg() %>% 
  set_engine("keras")

# Making first lm fit based on tidymodels, Y ~ DOSE
lm_mod1 <- linear_reg()

lm_fit1 <- 
  lm_mod1 %>% 
  fit(Y ~ DOSE, data = data_clean)

# Predictions: from ChatGPT
pred_lm1 <- predict(lm_fit1, data_clean) %>%
  bind_cols(data_clean)

# Metrics: from ChatGPT
metrics(pred_lm1, truth = Y, estimate = .pred)

# Creating table of coefficients
tidy(lm_fit1)

# Plotting first fit
tidy(lm_fit1) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))

```

```{r}
# Making second lm fit based on tidymodels, Y ~ DOSE + AGE + SEX + RACE + WT + HT
lm_mod2 <- linear_reg()

lm_fit2 <- 
  lm_mod2 %>% 
  fit(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = data_clean)

# Predictions: from ChatGPT
pred_lm2 <- predict(lm_fit2, data_clean) %>%
  bind_cols(data_clean)

# Metrics: from ChatGPT
metrics(pred_lm2, truth = Y, estimate = .pred)

# Creating table of coefficients
tidy(lm_fit2)

# Plotting second fit
tidy(lm_fit2) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

```{r}
#Loading engine for glm
logistic_reg() %>% 
  set_engine("glm")

# Making first glm fit based on tidymodels, Y ~ DOSE
log_mod1 <- logistic_reg() 

log_fit1 <- 
  log_mod1 %>% 
  fit(SEX ~ DOSE, data = data_clean)

# Predictions
pred_log1 <- predict(log_fit1, data_clean, type = "prob") %>%
  bind_cols(predict(log_fit1, data_clean)) %>%
  bind_cols(data_clean)

# Metrics
accuracy(pred_log1, truth = SEX, estimate = .pred_class)
roc_auc(pred_log1, truth = SEX, .pred_1)

# Creating table of coefficients
tidy(log_fit1)

# Plotting second fit
tidy(log_fit1) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))

```

```{r}
# Making second glm fit based on tidymodels, Y ~ DOSE + AGE + SEX + RACE + WT + HT
log_mod2 <- logistic_reg()  

log_fit2 <- 
  log_mod2 %>% 
  fit(SEX ~ DOSE + AGE + RACE + WT + HT, data = data_clean)

# Predictions
pred_log2 <- predict(log_fit2, data_clean, type = "prob") %>%
  bind_cols(predict(log_fit2, data_clean)) %>%
  bind_cols(data_clean)

# Metrics
accuracy(pred_log2, truth = SEX, estimate = .pred_class)
roc_auc(pred_log2, truth = SEX, .pred_1)

# Creating table of coefficients
tidy(log_fit2)

# Plotting second fit
tidy(log_fit2) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

#### The only worthly pattern I observed was the higher dose and higher 
#### The data was very skewed in certain categories like sex and race, Also there might be two possible age groups.
#### However it is hard to know whatg we are looking atg if I can not make a connection between them. 
#### Other than that, for the log model height it was linked to sex, which it kinda makes sense as height differs between male and females but no other predictor was significant


# MODULE 10
#### starts here

```{r}
data_clean2 <- data.frame("Y" = data_clean$Y, 
"DOSE" = data_clean$DOSE, 
"AGE" = data_clean$AGE, 
"SEX" = data_clean$SEX, 
"WT" = data_clean$WT, 
"HT" = data_clean$HT
)

rngseed = 1234
set.seed(rngseed)

# Put 3/4 of the data into the training set 
data_split <- initial_split(data_clean2, prop = 0.75)

# Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```


```{r}
new_model_null <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ 1, train_data)

tidy(new_model_null)

new_model1 <-  linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ DOSE, data = train_data)
tidy(new_model1)

new_model2 <-   linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ DOSE + AGE + SEX + WT + HT, data = train_data)
tidy(new_model2)

```


```{r}
#RMSE null
avg_resoonse <- mean(train_data$Y)
rmse_null_model <- sqrt(mean((train_data$Y - avg_resoonse)^2))
rmse_null_model

# RSME1
predict_new1 <- predict(new_model1, train_data) %>%
  bind_cols(train_data)
rmse_model1 <- rmse(predict_new1, truth = Y, estimate = .pred)
rmse_model1

#RMSE 2
predict_new2 <- predict(new_model2, train_data) %>%
  bind_cols(train_data)
rmse_model2 <- rmse(predict_new2, truth = Y, estimate = .pred)
rmse_model2

```


```{r}
set.seed(rngseed)
folds <- vfold_cv(data_clean2, v = 10)
folds

# First CV
new_model1_cv <- fit_resamples(
  linear_reg() %>% 
  set_engine("lm"),
  Y ~ DOSE,
  resamples = folds,
  metrics = metric_set(rmse)
)

# Second CV
new_model2_cv <- fit_resamples(
  linear_reg() %>% 
  set_engine("lm"),
  Y ~ DOSE + AGE + SEX + WT + HT,
  resamples = folds,
  metrics = metric_set(rmse)
)

# results
res_cv_model1 <- collect_metrics(new_model1_cv)
res_cv_model2 <- collect_metrics(new_model2_cv)

res_cv_model1
res_cv_model2
```

### Changes/improvements in 10A

```{r}
# Summary Table for 10A
model_comparison <- tibble(
  Model = c("Null Model", "Model 1 (Dose)", "Model 2 (Full)"),
  Train_RMSE = c(rmse_null_model, rmse_model1$.estimate, rmse_model2$.estimate),
  CV_RMSE = c(NA, res_cv_model1$mean, res_cv_model2$mean)
)
knitr::kable(model_comparison, digits = 2, caption = "Model Performance Comparison (10A)")
```

Based on the model comparison, Model 2 (Full Model) provided the best fit with an RMSE of 627 on the training data. The inclusion of physical characteristics, particularly Weight (WT), significantly improved the model's predictive power over the Dose-only model. The CV results confirm that this improvement is consistent across different folds.

# Code made by MR
#### (I contributed to other repo and waited but did not recieved the pull request from the other person, so I am pasting the code I contributed to them but I adapted it to mine)

```{r}
# This section added by MARCO REINA

#PART 2
#predicted values
data_clean2 <- data_clean2 %>%
  mutate(
    pred_null = predict(new_model_null, new_data = data_clean2, type = "numeric")$.pred,
    pred_mod1 = predict(new_model1, new_data = data_clean2, type = "numeric")$.pred,
    pred_mod2 = predict(new_model2, new_data = data_clean2, type = "numeric")$.pred
  )

```


```{r}
# Create a long-format data frame for plotting
pred_long_format <- data_clean2 %>%
  select(Y, pred_null, pred_mod1, pred_mod2) %>%
  pivot_longer(
    cols = starts_with("pred_"),
    names_to = "model",
    values_to = "prediction"
  ) %>%
  # Clean up model labels for nicer legend
  mutate(
    model = case_when(
      model == "pred_null" ~ "Null Model",
      model == "pred_mod1" ~ "Model 1 (DOSE)",
      model == "pred_mod2" ~ "Model 2 (Full)",
      TRUE ~ model
    )
  )
```

```{r}
# Plotting observed vs predicted
ggplot(pred_long_format, aes(x = Y, y = prediction, color = model, shape = model)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "gray50", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(limits = c(0, 6000)) + # englarged to 6000 because there were values outside the plot
  scale_y_continuous(limits = c(0, 6000)) + # englarged to 6000 because there were values outside the plot
  coord_fixed(ratio = 1) +  # Ensures 45° line appears at true 45°
  labs(
    title = "Observed vs. Predicted Values by Model",
    x = "Observed Values",
    y = "Predicted Values",
    color = "Model",
    shape = "Model"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  facet_wrap(~ factor(model,
  levels = c("Null Model", "Model 1 (DOSE)", "Model 2 (Full)")))
```


```{r}
#Plotting residuals
data_clean2$residuals_mod2 <- data_clean2$pred_mod2 - data_clean2$Y

ggplot(data_clean2, aes(x = Y, y = residuals_mod2)) + 
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color = "gray50", linewidth = 0.8, linetype = "dashed") +
  ylim(-2000, 2000) + #only 1 value left out to make it more visible
  theme_bw() 

```

```{r}
set.seed(rngseed) #seed set as rngseed

# Create 100 bootstrap samples from training data
dat_bs <- bootstraps(train_data, times = 100) 

num_obs <- nrow(train_data)
pred_bs <- matrix(NA, nrow = 100, ncol = num_obs)

for (i in 1:100) {
  # Get the i-th bootstrap sample
  dat_sample <- rsample::analysis(dat_bs$splits[[i]])
  
  # Fit the model (Model 2: all predictors) to this sample
  bs_fit <- linear_reg() |> 
    set_engine("lm") |> 
    fit(Y ~ DOSE + AGE + SEX + WT + HT, data = dat_sample)
  
  # Predict for the ORIGINAL training data
  # Use pull(.pred) to get a numeric vector instead of a dataframe
  pred_bs[i, ] <- predict(bs_fit, new_data = dat_sample) |> pull(.pred)
}

preds <- pred_bs |> apply(2, quantile, c(0.055, 0.5, 0.945)) |> t()
```

```{r}
# Get the original predictions from the Model 2 fit on training data

aug_train <- augment(new_model2, new_data = train_data)

# Combine with the bootstrap results (the 'preds' matrix from the previous step)
# This assumes 'preds' has columns for 5.5%, 50%, and 94.5% quantiles
plot_data <- aug_train |>
  mutate(
    bs_low = preds[, 1],
    bs_med = preds[, 2],
    bs_high = preds[, 3]
  )
```


```{r}
#Plotting the bootstrap results

ggplot(plot_data, aes(x = Y)) +
  # 45-degree reference line (Ideal model)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  # Bootstrap intervals (Upper and Lower bounds)
  geom_errorbar(aes(ymin = bs_low, ymax = bs_high), color = "skyblue", alpha = 0.5) +
  # Bootstrap median
  geom_point(aes(y = bs_med), color = "blue", size = 1, alpha = 0.6) +
  # Original point estimate (Original predictions)
  geom_point(aes(y = .pred), color = "black", shape = 1, size = 2) +
  # Formatting
  labs(
    x = "Observed Values",
    y = "Predicted Values (Point & Bootstrap Bounds)",
    title = "Observed vs. Predicted with Bootstrap Uncertainty"
  ) +
  # Ensure x and y axes are on the same scale
  coord_obs_pred() + 
  theme_bw()
```

#### MR: I copied and pasted this code submitted by ALEXANDRA TEJADA-STROP from a pull request, the merge had conflicts that could not be resolved. 

# This section added by ALEXANDRA TEJADA-STROP

```{r}
library(tidyverse)
library(tidymodels)

# Predicted versus observed plot for all 3 models on training data

# Null model predictions
null_preds_plot <- train_data %>%
  mutate(.pred = mean(train_data$Y), model = "Null model") %>%
  select(Y, .pred, model)

# Model 1 predictions
m1_preds_plot <- predict(new_model1, train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(model = "Model 1: DOSE only") %>%
  select(Y, .pred, model)

# Model 2 predictions
m2_preds_plot <- predict(new_model2, train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(model = "Model 2: All predictors") %>%
  select(Y, .pred, model)

# Combine all predictions
pred_plot_df <- bind_rows(null_preds_plot, m1_preds_plot, m2_preds_plot)

# Plot observed versus predicted
ggplot(pred_plot_df, aes(x = Y, y = .pred, color = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed versus predicted values on training data",
    x = "Observed Y",
    y = "Predicted Y",
    color = "Model"
  ) +
  theme_bw()
```

# Residual plot for Model 2
```{r}
resid_df <- predict(new_model2, train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(residual = .pred - Y)

ggplot(resid_df, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-2500, 2500)) +
  labs(
    title = "Residuals versus predicted values for Model 2",
    x = "Predicted Y",
    y = "Residuals"
  ) +
  theme_bw()
```

```{r}
# Bootstrap uncertainty for Model 2
set.seed(rngseed)
dat_bs <- bootstraps(train_data, times = 100)

pred_bs <- matrix(NA, nrow = 100, ncol = nrow(train_data))

for (i in 1:100) {
  dat_sample <- rsample::analysis(dat_bs$splits[[i]])
  
  bs_fit <- linear_reg() %>%
    set_engine("lm") %>%
    fit(Y ~ DOSE + AGE + SEX + WT + HT, data = dat_sample)
  
  pred_bs[i, ] <- predict(bs_fit, new_data = train_data)$.pred
}

preds <- apply(pred_bs, 2, quantile, c(0.055, 0.5, 0.945)) %>%
  t() %>%
  as.data.frame()

colnames(preds) <- c("lower", "median", "upper")

orig_preds <- predict(new_model2, train_data) %>%
  bind_cols(train_data %>% select(Y))

boot_plot_df <- orig_preds %>%
  bind_cols(preds)

ggplot(boot_plot_df, aes(x = Y)) +
  geom_point(aes(y = .pred), alpha = 0.7) +
  geom_point(aes(y = median), shape = 1, color = "blue") +
  geom_errorbar(aes(ymin = lower, ymax = upper), alpha = 0.3, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed versus predicted with bootstrap uncertainty",
    x = "Observed Y",
    y = "Predicted Y"
  ) +
  theme_bw()
```

