#Fitting exercise

```{r}
# Loading data
library(here)
data <- read.csv(here("fitting-exercise/", "Mavoglurant_A2121_nmpk.csv"))

str(data)

# Plotting DV ~ TIME
library(ggplot2)
ggplot(data, aes(y = DV, x = TIME)) + geom_point() + theme_bw()

ggplot(data, aes(y = DV, x = TIME, color = as.factor(DOSE))) + geom_point() + theme_bw()

ggplot(data, aes(y = DV, x = TIME, color = as.factor(DOSE))) + geom_point() + facet_wrap(~ DOSE) + theme_bw() + theme(strip.text = element_blank())
```

```{r}
# Removing OCC = 2
data_occ1 <- subset(data, OCC == 1)

# Removing time = 0
data_timefilter <-  subset(data_occ1, !(TIME == 0))

# Data only with time zero
data_timezero <-  subset(data_occ1, (TIME == 0))

library(dplyr)
# Summarizing - ChatGPT used for this code
sum_dv <- data_occ1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE))

# Merging datasets
merged_data <- left_join(data_timezero, sum_dv, by = "ID")
```

```{r}
# "Last" data cleaning
data_clean <- data.frame(Y  = merged_data$Y,
                        DOSE = merged_data$DOSE,
                        AGE = merged_data$AGE,
                        SEX = as.factor(merged_data$SEX),
                        RACE = as.factor(merged_data$RACE),
                        WT = merged_data$WT,
                        HT = merged_data$HT)

library(readr)
write_csv(data_clean, here("fitting-exercise/", "data_clean.csv"))
```

```{r}
# Plotting distributions for each one of the variables
hist(data_clean$Y) # Normally distirbuted, slight right skew
hist(data_clean$DOSE) # 3 doses used, will move to as.factor for certain plots
hist(data_clean$AGE) # few observation in 30-35 bin 
plot(data_clean$SEX) # sex 1 overrepresented
plot(data_clean$RACE) # Race 7 underrepresented
hist(data_clean$WT) # Normally distributed
hist(data_clean$HT) # Normally distributed
```

```{r}
# First round of plots
ggplot(data_clean, aes(y = Y, x = as.factor(DOSE))) + geom_boxplot() + theme_bw() # Hihgher dose, higher Y

ggplot(data_clean, aes(y = Y, x = AGE)) + geom_point() + theme_bw() # No clear pattern observed

ggplot(data_clean, aes(y = Y, x = SEX)) + geom_boxplot() + theme_bw() # Similar distributuion of SEX 1 and 2

ggplot(data_clean, aes(y = Y, x = RACE)) + geom_boxplot() + theme_bw() #Similar distribution for RACE 1, 2, 7, and 88

ggplot(data_clean, aes(y = Y, x = WT)) + geom_point() + theme_bw() # No clear pattern observed

ggplot(data_clean, aes(y = Y, x = HT)) + geom_point() + theme_bw() # No clear pattern observed

```

```{r}
# Second round of plots
ggplot(data_clean, aes(y = Y, x = as.factor(DOSE), levels =  SEX, fill = SEX)) + geom_boxplot() + theme_bw() # DOSE = 37.5 has wider distribution for SEX = 1

ggplot(data_clean, aes(y = Y, x = as.factor(DOSE), levels =  RACE, fill = RACE)) + geom_boxplot() + theme_bw() # DOSE = 37.5 has no representation of RACE = 7 and 88

ggplot(data_clean, aes(y = Y, x = WT, color = SEX)) + geom_point(size = 3) + theme_bw() + facet_wrap(~ SEX) + theme(strip.text = element_blank()) # SEX = 2 underrepresented as mentioned before, weight is less than SEX = 1

ggplot(data_clean, aes(y = Y, x = HT, color = RACE)) + geom_point(size = 3) + theme_bw() + facet_wrap(~ RACE) + theme(strip.text = element_blank()) # Only two observations of RACE = 7, very few observations of RACE = 88

ggplot(data_clean, aes(y = Y, x = WT, color = AGE, size = AGE)) + geom_point(alpha = 0.75) + theme_bw() # No clear pattern observed

ggplot(data_clean, aes(y = Y, x = HT, color = AGE, size = AGE)) + geom_point(alpha = 0.75) + theme_bw() # No clear pattern observed

```

```{r}
library(tidymodels)
library(dotwhisker)

#Loading engine for lm
linear_reg() %>% 
  set_engine("keras")

# Making first lm fit based on tidymodels, Y ~ DOSE
lm_mod1 <- linear_reg()

lm_fit1 <- 
  lm_mod1 %>% 
  fit(Y ~ DOSE, data = data_clean)

# Predictions: from ChatGPT
pred_lm1 <- predict(lm_fit1, data_clean) %>%
  bind_cols(data_clean)

# Metrics: from ChatGPT
metrics(pred_lm1, truth = Y, estimate = .pred)

# Creating table of coefficients
tidy(lm_fit1)

# Plotting first fit
tidy(lm_fit1) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))

```

```{r}
# Making second lm fit based on tidymodels, Y ~ DOSE + AGE + SEX + RACE + WT + HT
lm_mod2 <- linear_reg()

lm_fit2 <- 
  lm_mod2 %>% 
  fit(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = data_clean)

# Predictions: from ChatGPT
pred_lm2 <- predict(lm_fit2, data_clean) %>%
  bind_cols(data_clean)

# Metrics: from ChatGPT
metrics(pred_lm2, truth = Y, estimate = .pred)

# Creating table of coefficients
tidy(lm_fit2)

# Plotting second fit
tidy(lm_fit2) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

```{r}
#Loading engine for glm
logistic_reg() %>% 
  set_engine("glm")

# Making first glm fit based on tidymodels, Y ~ DOSE
log_mod1 <- logistic_reg() 

log_fit1 <- 
  log_mod1 %>% 
  fit(SEX ~ DOSE, data = data_clean)

# Predictions
pred_log1 <- predict(log_fit1, data_clean, type = "prob") %>%
  bind_cols(predict(log_fit1, data_clean)) %>%
  bind_cols(data_clean)

# Metrics
accuracy(pred_log1, truth = SEX, estimate = .pred_class)
roc_auc(pred_log1, truth = SEX, .pred_1)

# Creating table of coefficients
tidy(log_fit1)

# Plotting second fit
tidy(log_fit1) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))

```

```{r}
# Making second glm fit based on tidymodels, Y ~ DOSE + AGE + SEX + RACE + WT + HT
log_mod2 <- logistic_reg()  

log_fit2 <- 
  log_mod2 %>% 
  fit(SEX ~ DOSE + AGE + RACE + WT + HT, data = data_clean)

# Predictions
pred_log2 <- predict(log_fit2, data_clean, type = "prob") %>%
  bind_cols(predict(log_fit2, data_clean)) %>%
  bind_cols(data_clean)

# Metrics
accuracy(pred_log2, truth = SEX, estimate = .pred_class)
roc_auc(pred_log2, truth = SEX, .pred_1)

# Creating table of coefficients
tidy(log_fit2)

# Plotting second fit
tidy(log_fit2) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

#### The only worthly pattern I observed was the higher dose and higher 
#### The data was very skewed in certain categories like sex and race, Also there might be two possible age groups.
#### However it is hard to know whatg we are looking atg if I can not make a connection between them. 
#### Other than that, for the log model height it was linked to sex, which it kinda makes sense as height differs between male and females but no other predictor was significant


# MODULE 10
#### starts here

```{r}
data_clean2 <- data.frame("Y" = data_clean$Y, 
"DOSE" = data_clean$DOSE, 
"AGE" = data_clean$AGE, 
"SEX" = data_clean$SEX, 
"WT" = data_clean$WT, 
"HT" = data_clean$HT
)

rngseed = 1234
set.seed(rngseed)

# Put 3/4 of the data into the training set 
data_split <- initial_split(data_clean2, prop = 0.75)

# Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```


```{r}
new_model_null <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ 1, train_data)

tidy(new_model_null)

new_model1 <-  linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ DOSE, data = train_data)
tidy(new_model1)

new_model2 <-   linear_reg() %>%
  set_engine("lm") %>% 
  fit(Y ~ DOSE + AGE + SEX + WT + HT, data = train_data)
tidy(new_model2)

```


```{r}
#RMSE null
avg_resoonse <- mean(train_data$Y)
rmse_null_model <- sqrt(mean((train_data$Y - avg_resoonse)^2))
rmse_null_model

# RSME1
predict_new1 <- predict(new_model1, train_data) %>%
  bind_cols(train_data)
rmse_model1 <- rmse(predict_new1, truth = Y, estimate = .pred)
rmse_model1

#RMSE 2
predict_new2 <- predict(new_model2, train_data) %>%
  bind_cols(train_data)
rmse_model2 <- rmse(predict_new2, truth = Y, estimate = .pred)
rmse_model2

```


```{r}
set.seed(rngseed)
folds <- vfold_cv(data_clean2, v = 10)
folds

# First CV
new_model1_cv <- fit_resamples(
  linear_reg() %>% 
  set_engine("lm"),
  Y ~ DOSE,
  resamples = folds,
  metrics = metric_set(rmse)
)

# Second CV
new_model2_cv <- fit_resamples(
  linear_reg() %>% 
  set_engine("lm"),
  Y ~ DOSE + AGE + SEX + WT + HT,
  resamples = folds,
  metrics = metric_set(rmse)
)

# results
res_cv_model1 <- collect_metrics(new_model1_cv)
res_cv_model2 <- collect_metrics(new_model2_cv)

res_cv_model1
res_cv_model2
```
