# TidyTuesday exercise by MR
```{r}
#Loading the TidyTuesday library and data

library(tidytuesdayR)
library(dplyr)
library(tidymodels)
library(tidyverse)
library(here)
library(maps)
library(mapdata)

tuesdata <- tidytuesdayR::tt_load('2026-04-14')

#beaufort_scale <- tuesdata$beaufort_scale #Not necessary for my analysis
birds <- tuesdata$birds
#sea_states <- tuesdata$sea_states #Not necessary for my analysis
ships <- tuesdata$ships
```

# EDA

```{r}
# I took a look at the datasets in Positron and the Data Dictionary
# I am interested in understanding seasonal changes in birds sightings


# I decided to eliminate some of the columns that were mostly empty in both datasets (>50% empty)

birds$age = NULL
birds$wan_plumage_phase = NULL
birds$plumage_phase = NULL
birds$sex = NULL
birds$n_feeding = NULL
birds$n_sitting_on_water = NULL
birds$n_sitting_on_ice = NULL
birds$n_flying_past = NULL
birds$n_accompanying = NULL
birds$n_flying_past = NULL
birds$n_following_ship = NULL
birds$moulting = NULL

ships$direction = NULL
ships$wind_direction = NULL
ships$air_temperature = NULL
ships$pressure = NULL
ships$sea_surface_temperature = NULL
ships$depth = NULL
ships$time = NULL

# I also decided to remove "not interesting (for me)" columns

birds$bird_observation_id = NULL
birds$feeding = NULL
birds$sitting_on_water = NULL
birds$sitting_on_ice = NULL
birds$sitting_on_ship = NULL
birds$flying_past = NULL
birds$accompanying = NULL
birds$following_ship = NULL
birds$naturally_feeding = NULL
birds$in_hand = NULL

ships$speed = NULL
ships$direction = NULL
ships$cloud_cover = NULL
ships$precipitation = NULL
ships$wind_speed_class = NULL
ships$sea_state_class = NULL
ships$observer = NULL
ships$census_method = NULL
ships$hemisphere = NULL
ships$activity = NULL

```

### Let's merge both datasets

```{r}
merged_df <- left_join(birds, ships, by = "record_id")

saveRDS(merged_df, here("tidytuesday-exercise/tidytuesday_df.rds"))
```

### Let's make a some new features and summary table
```{r}
merged_df <- merged_df %>%
    mutate(
        aggr_spc = 1 + str_count(species_scientific_name, "/"),
        genus = word(species_scientific_name, 1),
        genus_code = substr(species_abbreviation, 1, 3),
        year  = year(date),
        month = month(date), 
        lat_bin = floor(latitude / 10) * 10, #creating larger bins for latitude and longitude
        lon_bin = floor(longitude / 10) * 10,
        count = as.numeric(count)
    ) %>%
        drop_na()

summary_genus <- merged_df %>%
    group_by(genus) %>% 
    summarise(
        count = n(), 
        .groups = "drop"
    ) %>%
    arrange(desc(count))

summary_genus %>%
    ggplot(aes(x = genus, y = count)) + geom_col() + coord_flip()
```

##### Seems like Diomeda is the genus with the majority of observations, I am going to focus on that. I would be interested to understand if thera are any trends I would like to focus on

```{r}
merged_df %>%
    filter(genus_code == "DIO") %>%
    ggplot(aes(x = year, y = count, fill = as.factor(aggr_spc))) + 
    geom_col()
```

#### It seems the major Genus observed it is observed as aggregated species (probably not fully distinguisable from afar to the osberver) 

#### Majority of the times Diomedea is either described into 2 species because both of them are very similar

#### Additionally sightings have two different time frames (before and after  1983)

### Let's see how frequent the sighting were in a map: all sightings from DIO

```{r}
# Get world map data
world_map <- map_data("world")

#All sightings of DIO
merged_df %>%
  filter(genus_code == "DIO") %>%
  group_by(lon_bin, lat_bin) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "gray50", size = 0.1) +
  geom_raster(aes(x = lon_bin, y = lat_bin, fill = count)) +
  scale_fill_viridis_c(option = "plasma", direction = -1, alpha = 0.5) +
  coord_quickmap(xlim = c(50, 180), ylim = c(-75, -25)) +
  theme_bw()

```

### Now let's look how it the two previous groups (before and after 1983) had different behaviors
```{r}
#Sightings before 1983
merged_df %>%
  filter(genus_code == "DIO", year <= 1983) %>%
  group_by(lon_bin, lat_bin) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "gray50", size = 0.1) +
  geom_raster(aes(x = lon_bin, y = lat_bin, fill = count)) +
  scale_fill_viridis_c(option = "plasma", direction = -1, alpha = 0.5) +
  coord_quickmap(xlim = c(50, 180), ylim = c(-75, -25)) + 
  theme_bw()

#Sightings after 1983
merged_df %>%
  filter(genus_code == "DIO", year > 1983) %>%
  group_by(lon_bin, lat_bin) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "gray50", size = 0.1) +
  geom_raster(aes(x = lon_bin, y = lat_bin, fill = count)) +
  scale_fill_viridis_c(option = "plasma", direction = -1, alpha = 0.5) +
  coord_quickmap(xlim = c(50, 180), ylim = c(-75, -25)) + 
  theme_bw()

```

#### Ok, from this information, I can see there are trends from each "wave" are differents (before and after 1983)

#### Let's keep looking, now within each of the "waves" per season.
```{r}
#Sightings before 1983 per season
merged_df %>%
  filter(genus_code == "DIO", year <= 1983) %>%
  group_by(lon_bin, lat_bin, season) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "gray50", size = 0.1) +
  geom_raster(aes(x = lon_bin, y = lat_bin, fill = count)) +
  scale_fill_viridis_c(option = "plasma", direction = -1, alpha = 0.5) +
  coord_quickmap(xlim = c(50, 180), ylim = c(-75, -25)) + 
  theme_bw() +
  facet_wrap(~season)

```

```{r}
#Sightings after 1983 per season
merged_df %>%
  filter(genus_code == "DIO", year > 1983) %>%
  group_by(lon_bin, lat_bin, season) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "gray50", size = 0.1) +
  geom_raster(aes(x = lon_bin, y = lat_bin, fill = count)) +
  scale_fill_viridis_c(option = "plasma", direction = -1, alpha = 0.5) +
  coord_quickmap(xlim = c(50, 180), ylim = c(-75, -25)) +
  theme_bw() +
  facet_wrap(~season)
```

#### It seems there is also a seasonal effect, we will take this into account for modeling

```{r}
#Let's create the dataset of interest
trends_df <- merged_df %>%
  filter(genus_code == "DIO") %>%
  group_by(lon_bin, lat_bin, season, year, month) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    wave = if_else(year >1983, 2, 1),
    geo = paste(lon_bin, lat_bin, sep = ", ")
  )

freq_df <- trends_df %>%
  group_by(geo) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(count))

freq_df
```

### From this information we can see the most "common" coordinates for sightings for DIO is 150, -40 with 7749 birds observed there.

### I am going to use this to measure "spread" I will use this as my base coordinate group and measure the distance between this high frequency coordinate to the actual

# ANALYSIS: 
### Hypothesis: The spread of DIO is greater in warmer months than in cold months/seasons. 
### Question: Is the spread getting smaller over the years?

```{r}
#Let's calculate distance

#Setting the main coordinate bins
lat_i <- -40
lon_i <- 150

# As the Earth is a globe, we need a special formula to calculate distance from A to B
library(geosphere)

trends_df <- trends_df %>%
  mutate(
    spreadkm = 1 + distHaversine(
      cbind(lon_i, lat_i), 
      cbind(lon_bin, lat_bin)
    )/1000,
    spread_rounded = round(spreadkm/100, digits = 0) * 100
  ) 

#Histograms to understand spread across both waves
trends_df %>%
  ggplot(aes(x = spread_rounded)) + 
  geom_histogram(binwidth = 1000) +
  facet_wrap(~ wave)

#Seasons in wave 1
trends_df %>%
  filter(wave == 1) %>%
  ggplot(aes(x = spread_rounded)) + 
  geom_histogram(binwidth = 1000) +
  facet_wrap(~ season)

#Seasons in wave 2
trends_df %>%
  filter(wave == 2) %>%
  ggplot(aes(x = spread_rounded)) + 
  geom_histogram(binwidth = 1000) +
  facet_wrap(~ season)
```

### From this information we can see that wave 1 have a wider spread in the autumn and summer. Wave 2 seems more compact across seasons, spring has the wider spread. Which it confirms what we observed on the maps.
### All of the trends are right skewed regardless of the season or wave.

# MODELING
# Now let's see if we can predict spread based on temporal predictors
```{r}
trends_df %>%
  mutate(
    year = as.factor(year),
    month = as.factor(month)
  )

trends_df$spread_log2 <- log(trends_df$spreadkm, base = exp(2)) + 1

#Creating a null linear model
lm_null <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ 1, data = trends_df)

#Including all variables and a stepwise approach (reverse)
lm_fit1 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ wave + season + year + month + count + lat_bin + lon_bin, data = trends_df)

lm_fit2 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ wave + season + year + month + lat_bin + lon_bin, data = trends_df)

lm_fit3 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ wave + season + year + lat_bin + lon_bin, data = trends_df)

lm_fit4 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ wave + season + lat_bin + lon_bin, data = trends_df)

lm_fit5 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ season, data = trends_df)

lm_fit6 <- linear_reg() %>%
  set_engine("lm") %>% 
  fit(spreadkm ~ wave, data = trends_df)

```

### Let's get the predictions
```{r}
#Predidiction and differences for each fit
trends_df <- trends_df %>%
  mutate(
    pred0 = predict(lm_null, new_data = trends_df, type = "numeric")$.pred,
    pred1 = predict(lm_fit1, new_data = trends_df, type = "numeric")$.pred,
    pred2 = predict(lm_fit2, new_data = trends_df, type = "numeric")$.pred,
    pred3 = predict(lm_fit3, new_data = trends_df, type = "numeric")$.pred,
    pred4 = predict(lm_fit4, new_data = trends_df, type = "numeric")$.pred,
    pred5 = predict(lm_fit5, new_data = trends_df, type = "numeric")$.pred,
    pred6 = predict(lm_fit6, new_data = trends_df, type = "numeric")$.pred
  ) %>%
    mutate(
    res0 = pred0 - spread_log2,
    res1 = pred1 - spread_log2,
    res2 = pred2 - spread_log2,
    res3 = pred3 - spread_log2,
    res4 = pred4 - spread_log2,
    res5 = pred5 - spread_log2,
    res6 = pred6 - spread_log2
  )
```

### Checking for predicted vs actual

```{r}
#Null model
ggplot(trends_df, aes(x = spreadkm, y = pred0)) +
    geom_point(size = 2) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = pred1)) +
    geom_point(size = 2) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = pred2)) +
    geom_point(size = 2) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = pred3)) +
    geom_point(size = 2) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = pred4)) +
    geom_point(size = 2) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = pred5)) +
    geom_point(size = 2) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = pred6)) +
    geom_point(size = 2) +
    theme_bw()
```

### Checking the residuals

```{r}
ggplot(trends_df, aes(x = spreadkm, y = res0)) +
    geom_point(size = 1, alpha = 0.5) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = res1)) +
    geom_point(size = 1, alpha = 0.5) +
    theme_bw()

ggplot(trends_df, aes(x = spreadkm, y = res2)) +
    geom_point(size = 1, alpha = 0.5) +
    theme_bw()
```



