# Description of the data
The BEAM (Bacteria, Enterics, Amoeba, and Mycotics) Dashboard is an interactive tool to access and visualize data from the System for Enteric Disease Response, Investigation, and Coordination (SEDRIC). The BEAM Dashboard provides timely data on pathogen trends and serotype details to inform work to prevent illnesses from food and animal contact.
Last updated: January 20, 2026
Data provided by the Division of Foodborne, Waterborne, and Environmental Diseases (DFWED)

License	Public Domain U.S. Government
Source Link	https://www.usa.gov/government-works

# Loading data
```{r}
library(here)
cdcdata <- read.csv(here("cdcdata-exercise", "BEAM_Dashboard_-_Report_Data_20260211.csv"))
```

# Data exploration
```{r}
# get an overview of data structure
str(cdcdata)

# get a summary of data
summary(cdcdata)
```

# Data cleaning
```{r}
#step 1 removing mostly empty columns
cdcdata_1 <- subset(cdcdata, select = -c(Outbreak.associated.isolates, New.multistate.outbreaks, New.multistate.outbreaks...US, X..Isolates.with.clinically.important.antimicrobial.resistance, Number.of.sequenced.isolates.analyzed.by.NARMS))

#step 2 dropping NA values
cdcdata_2 <- na.omit(cdcdata_1)
```

# Data re-exploration
```{r}
# get an overview of data structure
str(cdcdata_2)

# get a summary of data
summary(cdcdata_2)
```

# Number of isolates per year
```{r}
library(ggplot2)

ggplot(cdcdata_2, aes(x = Year, fill = Pathogen)) + geom_bar() + theme_bw()

ggplot(cdcdata_2, aes(x = Year)) + geom_bar(fill ="lightblue3") + theme_bw() + facet_wrap( ~ Pathogen, scales = "free")
``` 

# Trends by month
```{r}
ggplot(cdcdata_2, aes(x = Month, color = Year)) + geom_line(stat = "count") + theme_bw() + facet_wrap( ~ Pathogen, scales = "free")
```

# Comparing pathogens per state
```{r}
ggplot(cdcdata_2, aes(x = State, fill = Pathogen)) + geom_bar() + theme_bw() + facet_wrap(~ Pathogen, ncol = 1, scales = "free") + theme(axis.text.x = element_text(angle = 90), strip.text.x = element_blank(), legend.position = "top") 
```

# Source type
```{r}
ggplot(cdcdata_2, aes(x = Year, fill = Pathogen)) + geom_bar() + theme_bw() + facet_grid(Source.Type ~ Pathogen, scales = "free") + theme(legend.position = "none")
```
